---
title: "BIO532 Introduction to R"
author: "Michal L Collyer"
date: "9/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BIO532 R INTRODUCTION, just some basics for using R

First thing to note, if one wants to add notes to an R script one must use a # to type the note. DO THIS OFTEN!!!!  IT HELPS OTHERS UNDERSTAND YOUR INTENTION.  It also reminds you what you were attempting to do later.  Finally, you might have to submit scripts as homework.  If you have trouble, it's easier to communicate what you were attempting to do with annotated scripts.

### Typing in data or data files

```{r, include=TRUE, echo = TRUE}
a = 4
a

a <- 4
a
```

Note: R is an open-source programming language that was modeled after the S programming language (yes, the name is a bit of a play on the name S, but the original authors had the first names Ross and Robert).  The creator of the S programming language is a member of the R Core Development team (so no competition). R makes a point to work with all S programming.  In S, one must do the following, which also works perfectly fine in R

Note that spaces are not needed.  Note that the arrow assigns an object a value, a string of values, or a function. Note that in R, as we will see later, an actual equality might need `==` (two equal signs) to differentiate it from an assignment.  Therefore, I recommend using arrows instead of equal sign.

### Other types of assignments

```{r, include=TRUE, echo=TRUE}

b <- c(4, 2, 5) # c means concatenate or combine
c <- array(1:5)
c <- 1:5
d <- array(1, 10)
e <- array(1:3, 15)
f <- rep(seq(1, 4, 1), 5) # seq means sequence (from 1 to 4 by 1s); rep = repeat (5 times)
g <- c("I", "LOVE","R",4,"STATISTICS")
h <- c("I love R for statistics")

a
b
c
d
e
f
g
h

```

If one wants to assign a few variables they can type it all in and assemble a data frame; e.g.,

```{r, include=TRUE, echo=TRUE}

x = c(10, 11, 14, 15, 9, 12, 8)
y = c(100, 100, 110, 100, 120, 140, 110)
z = factor(c(rep("A", 4),rep("B", 3))) # factor means a nominal/categorical variable

x
y
z

Y = data.frame(x,y,z)
Y # each row is a subject with three values for three variables/variates
```

Data Frames are the most common way to deal with data for statistical analysis.  They are matrices, but with the specific designation that columns are variables

### Reading in files
One will do this a lot, if one has data, say, in an Excel file or other type of spreadsheet.  IT IS BEST TO SAVE EXCEL FILES AS TEXT DOCUMENTS.  FOR EXAMPLE, "save as", THEN CHOOSE .csv or .txt.  There are different formats to use, but .csv (comma separated values) files work regularly and do not require knowing the difference between a space and a tab (values in text files can be separated or 'delimited' by commas, spaces, or tabs). **At this point, please download the CSV file, lowbwt.csv, from Brightspace/Resources/Data Files.** Make sure you know the location of the file on your computer.  lowbwt.csv is a set of data for babies born with low birthweight (from Pagano and Gauvreau, Biostatistics, 1992).

The simple way (if you like clicking buttons): 

+ Look over to the upper-right of R-Studio,
+ Click on "Environment" tab, then "Import Dataset".
+ Should be sort of obvious what to do...

The purer way (if you like command line text)" 

+ First, change the directory in R to match that directory 
+ (Session/Set Working Directory/Choose Location)
  + Go to the "Session" pull-down menu
  + Go to "Set Working Directory"
  + Find the directory where the spreadsheet is saved.
  
```{r, include = FALSE, echo = FALSE}
lowbwt <- read.csv("https://raw.githubusercontent.com/mlcollyer/chatham.bio532/master/Data/lowbwt.csv")
```

  
```{r, include=TRUE, echo=TRUE}
# lowbwt = read.csv("low.bwt.csv",header=T) Remove the # when you do this for real...
```

Look at the data in the environment tab (click on it in the environment tab - opens a spreadsheet).  Or just do this:

```{r}
lowbwt
```

or this:

```{r}
View(lowbwt)
```

No matter how the data were opened, look at the console.  `read.csv` is a function.  It performs a task.  If you ever need help using a function use `?` followed by the function; e.g., 

```{r}
? read.csv
? factor
? seq
```

Note that any spreadsheet in Excel can be saved as a text file.  One must do this to read the file into R.  Other stats programs might be able to read Excel files.  There used to be a package for doing this in R but I believe it had trouble keeping up with Microsoft updates.  Using ext files is universal.  JUST MAKE TEXT FILES!  Saves some headaches.

NOTE, When one wants to read a file from a URL directly into R....

```{r, include = TRUE, echo = FALSE}
lowbwt <- read.csv("https://raw.githubusercontent.com/mlcollyer/chatham.bio532/master/Data/lowbwt.csv")
```

### Playing with the data

```{r}
dim(lowbwt) # provides the dimensions of the "data frame"
str(lowbwt) # an overview of the structure of the object, lowbwt

lowbwt[33,2] # provides the value in the 33rd row and 2nd column
lowbwt[33,] # provides the entire 33rd row
lowbwt[,2] # provides the entire 2nd column

lowbwt[1:10,] # provides the first 10 rows
lowbwt[-(3:20),] # provides the entire data frame, minus the third through twentieth rows
```

When objects have "sublevels", the symbol,'$',is used to strictly pay attention to a sublevel; e.g.,

```{r}
str(lowbwt)
```

Make certain variables are turned into factors

```{r}
lowbwt$sex <- factor(lowbwt$sex)
lowbwt$tox <- factor(lowbwt$tox)
lowbwt$grmhem <- factor(lowbwt$grmhem)

str(lowbwt)

lowbwt$sbp
lowbwt$apgar5
```


### COMMON SUMMARY FUNCTIONS

+ summary, mean, median, quantile, min, max, var, sd, IQR,
+ plot (generic), barplot, boxplot, hist, rug, table

Remember, for R, a qualitative variable is a "factor"; quantitative variables can be integer (discrete) or numeric (continuous)

### Quick summaries

```{r}
summary(lowbwt) 
```

Quartiles...  (Note difference between quantile and quartile)

```{r}
quantile(lowbwt$sbp, c(0.25, 0.50, 0.75)) # Quartiles are quantiles; not the other way around

quantile(lowbwt$sbp, 0.37) # 37th percentile (quantiles also called percentiles)
```

### Histogram

```{r}
hist(lowbwt$sbp)
```

Need some help?

```{r}
?hist

hist(lowbwt$sbp, right = FALSE, 
     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90))

# let's get fancier...

hist(lowbwt$sbp, right = FALSE, 
     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90), 
     col = "yellow")

hist(lowbwt$sbp, right = FALSE, 
     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90), 
     col = "yellow",
     xlab = "Systolic Blood Pressure, mmHG",
     ylab = "Frequency")
```

### More plots

```{r}
boxplot(lowbwt$sbp, ylab = "Systolic Blood Pressure, mm Hg")

boxplot(lowbwt$sbp, horizontal = TRUE, 
        xlab = "Systolic Blood Pressure, mm Hg")


boxplot(lowbwt$sbp, horizontal = TRUE, 
        xlab = "Systolic Blood Pressure, mm Hg",
        notch = TRUE,
        col = "dark red")

?boxplot # If help is needed
```

### Dispersion

```{r}
var(lowbwt$sbp) # sample variance
sd(lowbwt$sbp) # sample SD
IQR(lowbwt$sbp)
```


# Summary
There are so many things we can do with R but it can also be overwhelming.  Future exercises will introduce new functions to use.  Do not worry about how much you understand right now.  Here are some goals for this session:

1. Be able to read in data frames
2. Understand what a function is/does; e.g., var( )
3. Understand how to find help; e.g., ? var
4. Get comfortable moving around in R Studio

