---
title: "BIO532 Linear Algebra in R"
author: "Michal L Collyer"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up some matrices comprising vectors

```{r}
y1 <- c(2, 3, 2, 5, 6, 8) # variable 1 with 6 observations
y2 <- c(-1, -2, 0, 0, 1, -1) # variable 2 with 6 observations
Y <- as.matrix(data.frame(y1 = y1, y2 = y2))
rownames (Y) <- paste("obs", 1:6, sep = ".") # giving our observations some names


x1 <- rep(1, 6) # variable x1 with 6 observations
x2 <- c(rep(0, 3), rep(1,3)) # variable x2 with 6 observations
X <- as.matrix(data.frame(x1 = x1, x2 = x2))
rownames(X) <- rownames(Y)

X
Y
```

# Inner-products.  

Note how many are possible!

```{r}
crossprod(x1)
crossprod(x2)
crossprod(x1, x2)
crossprod(X)

crossprod(y1)
crossprod(y2)
crossprod(y1, y2)
crossprod(Y)

crossprod(x1, y1)
crossprod(x1, y2)
crossprod(x2, y1)
crossprod(x2, y2)
crossprod(X, Y)
crossprod(Y, X)

```

# Invert a matrix

```{r, error = TRUE}

solve(X) # does not work

```

```{r}
solve(crossprod(X))
```


# Prelude to linear models

Just some example but this will be covered in more detail in the next script.

```{r}
solve(crossprod(x1)) %*% crossprod(x1, y1) # mean
solve(crossprod(X)) %*% crossprod(X, y1) # group means
solve(crossprod(X)) %*% crossprod(X, Y) # group means for two variables
```


# Think about how long this would take to do this "by hand"

One should never forget or fail to appreciate just how pwerful computers are.  Imagine this course 40 years ago, and being asked to do almost all operations with just a pencil and basic calculator or adding machine!

```{r}
# 10,000 x 100 matrices of random values
Y <- matrix(rnorm(1000000), nrow = 10000, ncol = 100) 
X <- matrix(rnorm(1000000), nrow = 10000, ncol = 100) 

XtY <- crossprod(X, Y) # transpose of X times Y
dim(XtY)
XYt <- tcrossprod(X, Y) # X times the transpose of Y
dim(XYt)

# the first does a 100 x 100 calculation of inner products, each 10,000 values
# the second does a 10,000 x 10,000 calculation of inner products, each 100 values

system.time(crossprod(X, Y))
system.time(tcrossprod(X, Y))

# The difference is staggering but realize that 10,000 x 10,000 = 100,000,000 operations!
# yet the computer did it in probably less than 1/2 minute!
```




