---
title: "BIO532 Single-factor ANOVA"
author: "Michal L Collyer"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# The same examples as in Chapter 6

IMPORTANT! BEFORE GOING FURTHER, DO THESE TWO THINGS in R"

install.packages("devtools") (Do this just this once)

devtools::install_github("mlcollyer/RRPP", 
                         build_vignettes = TRUE) (Do this every script)
                         
The reason for this is that `RRPP` is a packaged published on CRAN but updating it on CRAN takes a long time.  Since the instructor is also the `RRPP` author, updates can be made instantly on `Github`.  Installing directly from `Githuib` every `R` session means having the latest updates.


# Read in the calluna data
Either use the Import Dataset tab or change working directory to where the `calluna.csv` file is found and run command below.  (To change the directory in R Studio: Session --> Set Working Directory --> Choose location.  A pop-up window will appear.)

```{r, include=TRUE, echo=TRUE}
calluna <- read.csv("calluna.csv", 
                       header = TRUE, 
                       stringsAsFactors = TRUE)

str(calluna) # or view in Environment tab

calluna <- calluna[c("treatment", "cn.ratio")] # eliminate unneeded columns
calluna <- na.omit(calluna) # Eliminate rows with NA as values
str(calluna)
```


# Plot
```{r, include=TRUE, echo=TRUE}
boxplot(cn.ratio ~ treatment, data = calluna, 
        border = 1:4, col = "gray")
stripchart(cn.ratio ~ treatment, data = calluna, 
           pch = 19, col = 1:4, vertical = TRUE,
           add = TRUE)
```

# State null and alternative hypotheses:
$$H_0: \sigma^2_{Treatment|\mu} = 0$$
$$H_a: \sigma^2_{Treatment|\mu} > 0$$

# Confidence level = 95% ($\alpha = 0.05$)

# Fit linear model

```{r, include=TRUE, echo=TRUE}
library(RRPP) # open RRPP library

fit1 <- lm.rrpp(cn.ratio ~ treatment, 
                data = calluna, 
                iter = 9999, 
                print.progress = FALSE)


```

# Linear model parts
```{r, include=TRUE, echo=TRUE}

model.matrix(fit1)
coef(fit1)
fitted(fit1)
residuals(fit1)

```

# Residual diagnostics
```{r, include=TRUE, echo=TRUE}
plot(fit1, which = c(1, 2))
```

# Perform ANOVA
```{r, include=TRUE, echo=TRUE}
anova(fit1)
```

# Conclusion?

### Reject $H_0$; `treatment` important term in model; $Z$-score suggests it is a large effect

# We are not done !!!

We have multiple potential reasons for why this null hypothesis is rejected.

```{r, include=TRUE, echo=TRUE}
PW <- pairwise(fit1, groups = calluna$treatment)
summary(PW, confidence = 0.95, show.vectors = TRUE)
```
# Conclusion?

In terms of means,
+ D no different than Control
+ N > Control
+ D + N > Control
+ D + N > N?  (Close but cannot reject H0)  Can say the result is "marginally" significant, but this biological interpretation must be germane.
