---
title: "BIO532 ANOVA"
author: "Michal L Collyer"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# The same examples as in Chapter 5

IMPORTANT! BEFORE GOING FURTHER, DO THESE TWO THINGS in R"

install.packages("devtools") (Do this just this once)

devtools::install_github("mlcollyer/RRPP", 
                         build_vignettes = TRUE) (Do this every script)
                         
The reason for this is that `RRPP` is a packaged published on CRAN but updating it on CRAN takes a long time.  Since the instructor is also the `RRPP` author, updates can be made instantly on `Github`.  Installing directly from `Githuib` every `R` session means having the latest updates.


# Read in the pupfish MR data
Either use the Import Dataset tab or change working directory to where the `pupfish.mr.csv` file is found and run command below.  (To change the directory in R Studio: Session --> Set Working Directory --> Choose location.  A pop-up window will appear.)

```{r, include=TRUE, echo=TRUE}
pupfish.mr <- read.csv("pupfish.mr.csv", 
                       header = TRUE, 
                       stringsAsFactors = TRUE)

# Add a variable to the data frame
pupfish.mr$SMR <- pupfish.mr$MR / pupfish.mr$Mass

str(pupfish.mr) # or view in Environment tab
```

# Add a variable to the data frame

```{r, include=TRUE, echo=TRUE}
```
pupfish.mr$SMR <- pupfish.mr$MR / pupfish.mr$Mass

# Plot
```{r, include=TRUE, echo=TRUE}
boxplot(SMR ~ Treatment, data = pupfish.mr, 
        border = 1:2, col = "gray", notch = TRUE)
        
stripchart(SMR ~ Treatment, data = pupfish.mr, 
           pch = 19, col = 1:2, vertical = TRUE,
           add = TRUE)
```

# State null and alternative hypotheses:
$$H_0: \sigma^2_{Treatment|\mu} = 0$$
$$H_a: \sigma^2_{Treatment|\mu} > 0$$

# Confidence level = 95% ($\alpha = 0.05$)

# Fit linear model

```{r, include=TRUE, echo=TRUE}
library(RRPP) # open RRPP library

fit1 <- lm.rrpp(SMR ~ Treatment, 
                data = pupfish.mr, 
                iter = 9, 
                print.progress = FALSE)

```

# Linear model parts
```{r, include=TRUE, echo=TRUE}
model.matrix(fit1)
coef(fit1)
fitted(fit1)
residuals(fit1)
```

# Residual diagnostics
```{r, include=TRUE, echo=TRUE}
plot(fit1, which = c(1, 2))
```

# Perform ANOVA
```{r, include=TRUE, echo=TRUE}
anova(fit1)
```

# How does RRPP change things?
```{r, include=TRUE, echo=TRUE}
fit2 <- lm.rrpp(SMR ~ Treatment, 
                data = pupfish.mr, 
                iter = 99, 
                print.progress = FALSE)
fit3 <- lm.rrpp(SMR ~ Treatment, 
                data = pupfish.mr, 
                iter = 999, 
                print.progress = FALSE)
fit4 <- lm.rrpp(SMR ~ Treatment, 
                data = pupfish.mr, 
                iter = 9999, 
                print.progress = FALSE)
fit5 <- lm.rrpp(SMR ~ Treatment, 
                data = pupfish.mr, 
                iter = 99999, 
                print.progress = FALSE)

anova(fit1)
anova(fit2)
anova(fit3)
anova(fit4)
anova(fit5)
```           

# Conclusion?

### Reject $H_0$; `Treatment` important term in model; $Z$-score suggests it is a large effect